---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---



## Loading and preprocessing the data

```{r}
options(scipen=999)
library(dplyr)
data <- read.csv(file = "activity.csv", header = TRUE)
```


## What is mean total number of steps taken per day?
```{r}
dailysteps <- data %>% group_by(date) %>% summarise(sum(steps, na.rm = TRUE))
names(dailysteps) = c("date", "daily.steps")
hist(dailysteps$daily.steps, breaks = 15, xlab = "daily steps", main = "Histogram of the average daily steps")
```

```{r}
average <- round(mean(dailysteps$daily.steps), digits = 0)
average
```
The mean total number of steps taken per day is `r average`

```{r}
median <- median(dailysteps$daily.steps)
median
```
The mean total number of steps taken per day is `r median`

## What is the average daily activity pattern?
```{r}
intervallsteps <- data %>% group_by(interval) %>% summarise(mean(steps, na.rm = TRUE))
names(intervallsteps) = c("interval", "av.daily.steps")
plot(intervallsteps$interval, intervallsteps$av.daily.steps, type = "l", xlab = "time interval", ylab = "average number of steps")
```

```{r}
maxtime <- intervallsteps[which.max(intervallsteps$av.daily.steps), 1]
maxtime
```
The intervall at `r maxtime` has the maximum average daily steps

## Imputing missing values
We're going to fill the missing values with the mean of the correspondent interval

```{r}
newdata <- mutate(data, missing.data = (is.na(steps)| is.na(date)))
newdata <- full_join(newdata, intervallsteps, by = "interval")
newdata[is.na(newdata)] <- 0
newdata <- mutate(newdata, new.steps = steps + (missing.data * av.daily.steps))
newdata <- select(newdata, date, interval, new.steps)

newdailysteps <- newdata %>% group_by(date) %>% summarise(sum(new.steps, na.rm = TRUE))
names(newdailysteps) = c("date", "daily.steps")
hist(newdailysteps$daily.steps, breaks = 15, xlab = "daily steps", main = "Histogram of the average daily steps after imputing the missing values")
```

We can compare the 2 histograms

```{r}
par(mfrow=c(1,2))
hist(dailysteps$daily.steps, breaks = 15, xlab = "daily steps", main = "omitt NA")
hist(newdailysteps$daily.steps, breaks = 15, xlab = "daily steps", main = "Replace NA")
```

As for the numerical values:
```{r}
newaverage <- round(mean(newdailysteps$daily.steps), digits = 0)
newaverage
```

```{r}
newmedian <- median(newdailysteps$daily.steps)
newmedian
```

```{r}
avdiff <- newaverage - average
meandiff <- newmedian - median

```
The new average is `r newaverage` which differs by `r avdiff` to before    
The new median is `r newmedian` which differs by `r meandiff` to before  

As we can see the mean differs quite a lot the median in this example not at all. This is because the median is much more robust than the mean. But in general one has to be careful when replacing NA-Values, because it can change the data a lot.


## Are there differences in activity patterns between weekdays and weekends?
```{r}
newdata <- mutate(newdata, weekday = weekdays(as.Date(date, format = "%Y-%m-%d")))

weekends <- filter(newdata, weekday == "Saturday" | weekday =="Sunday")
noweekends <- filter(newdata, weekday =="Monday" | weekday == "Tuesday" | weekday=="Wednesday" | weekday == "Thursday" | weekday == "Friday")

intervallstepsweekend <- weekends %>% group_by(interval) %>% summarise(mean(new.steps))
names(intervallstepsweekend) = c("interval", "av.daily.steps")

intervallstepsweekday <- noweekends %>% group_by(interval) %>% summarise(mean(new.steps))
names(intervallstepsweekday) = c("interval", "av.daily.steps")

```

Now we can make a panel plot comparing the two timeseries:

```{r}
par(mfrow=c(2,1), mar = c(2,4,1.5,1.5))
plot(intervallstepsweekend$interval, intervallstepsweekend$av.daily.steps, type = "l", xlab = "time interval", ylab = "average number of steps")
legend("topleft", legend = "weekend")

plot(intervallstepsweekday$interval, intervallstepsweekday$av.daily.steps, type = "l", xlab = "time interval", ylab = "average number of steps", )
legend("topleft", legend = "weekday")

```
